---
title: "Interim project"
output: html_notebook
---

Members: Daryna Horetska, Yaryna Pechenenko, Andrii Hovorov
Our dataset:
Multi-Resistance Antibiotic Susceptibility
https://www.kaggle.com/datasets/adilimadeddinehosni/multi-resistance-antibiotic-susceptibility
Our goal: to analyse bacteria resistance to different antibiotic depending on age, previous diseases, gender and etc.

Cleaning data:

```{r}
library(tidyr)

dataset_readed <- read.csv("Bacteria_dataset_Multiresictance.csv")

dataset_readed[dataset_readed == "?"] <- NA
dataset_readed[dataset_readed == "missing"] <- NA
dataset_readed[dataset_readed == "unknown"] <- NA
dataset_readed[dataset_readed == "error"] <- NA
dataset_readed[dataset_readed == "???"] <- NA
dataset_readed[dataset_readed == "??"] <- NA
dataset_readed[dataset_readed == ""] <- NA
dataset_readed[dataset_readed == "None"] <- NA


dataset_readed <- dataset_readed %>%
  separate(`age.gender`, into = c("age", "gender"), sep = "/", remove = FALSE)
```


```{r}
dataset_readed$age <- as.numeric(dataset_readed$age)

cat("Descriptive analysis of ages\n")
summary(dataset_readed$age)
cat("Infection frequency")
table(dataset_readed$Infection_Freq, useNA = "ifany")

```
Hypothesis:

1. Association between bacteria type and resistance
H0 -  There is no association between the bacterial species  and resistance to a specific antibiotic
H1 - There is an association between the bacterial species and resistance to the specific antibiotic

2. Impact of risk factors on resistance
H0 - There is no association between having the risk factor and being infected with a resistant strain. The probability of resistance is the same for patients with and without the risk factor
H1 - There is an association between having the risk factor and being infected with a resistant strain. The probability of resistance is different for patients with and without the risk factor

3. Effect of patient age on resistance
H0 - The mean age of patients with a resistant infection is equal to the mean age of patients with a susceptible infection
H1 - The mean age of patients with a resistant infection is not equal to the mean age of patients with a susceptible infection

4. Co-resistance between antibiotics
H0 - Resistance to one antibiotic is independent of resistance to the other
H1 - Resistance to one antibiotic is dependent of resistance to the other

